An Intrinsically-Typed Solution for the\\ List-Machine Benchmark

### 2021-03-10 papers/COLA/paper.tex
Finished the paper first version

In this work, we extend our previous SBLP 2020 paper by including support to
indirect jumps, as specified in the version 2.0 of the list
machine benchmark~\cite{AppelDL12}. We provide detailed explanations of the needed
changes on the machine syntax, type system and interpreter to accomodate this extension, which
allows the handling of more complex control flow, like the call/return model of function calls.

More specifically, our contributions are:

\begin{itemize}
  \item We show how all the details of the list-machine type system
        can be encoded in a dependently-typed syntax setting, which avoids, by construction,
        the presence of stuck states in its definitional interpreter.
  \item We provide a provably correct implementation for testing the subtyping
        relation and to calculate the least common supertype of two input
        types for the machine registers.
  \item We provide a correct-by-construction implementation of a type-checker for
        list-machine programs. Our type-checker produces, as a result, an intrinsically-typed
        representation of the machine code.
  \item We compare the results of an intrinsic approach encoded in Agda with extrinsic
        formalizations encoded in Twelf and Coq. We show that such intrinsic encoding avoids
        unnecessary repetitions and provides some properties for free.
  \item We provide a detailed discussion on the necessary type system and interpreter modifications
        to support indirect jumps.
\end{itemize}

@@

Appel's list machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics. We stick to our initial
proposal based on dependently typed syntax and definitional interpreters.

@@

\paragraph{Benchmarks for PL mechanization}{
The importance of benchmarks is recognized in several areas in computer science. A relevant example is the
TPTP library for the theorem proving community~\cite{Sut17}. In the context of the programming languages community,
we have the POPLMark challenge~\cite{Aydemir05}, which was developed by a renowned group of programming language researchers aiming
at the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
was to motivate authors to formalize all of their theorems using such tools. Since the focus of the POPLMark challenge
was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
machine benchmark was proposed by~\citet{Appel07} as an exercise in formalizing results that interest
compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
solution to this benchmark using an intrinsically-typed approach in the Agda programming language. Representation of binding syntax
was the subject of~\citet{FeltyMP18} which proposed a set of problems and research questions for tools
that use the high-order abstract syntax approach for name binding. Since the list-machine benchmark definition avoids
name binding issues because they are orthogonal to most of compiler related proofs~\cite{Appel07}, we just ensure
the correct manipulation of names by following the traditional \emph{de Bruijn} approach. Finally, a recent problem set
was proposed by~\citet{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
of logical relation arguments, like strong normalization theorems.}

\paragraph{Definitional interpreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
Recently, the interest on such interpreters was revitalized by~\citet{Amin17}, which used definitional
interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
Unlike our work, Amin and Rompf's formalizations do not use intrinsically-typed representations of their syntax, what cluttered
their formalization with the need of dealing with ``stuck states'' in the semantics. \citet{Poulsen18} described how
definitional interpreters for imperative languages can be much concisely implemented by using intrinsically-typed syntax and a
library for name binding using \emph{scope graphs}, which greatly simplifies the treatment of complex
binding structures. Since the list-machine benchmark was designed to address other problems than binding, our representation
using \emph{de Bruijn} indices was sufficient to implement the desired type-checker and interpreter.
Other recent application of definitional interpreters was proposed by~\citet{Rouvoet20}. The main contribution
of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
linear typing features in the interpreter, the authors have implemented monads based on a separation algebra which they
named \emph{Market}, supporting the main operations for accessing and updating the store used by the interpreter.
Another work using intrinsically-typed syntax for resource control was developed by~\citet{Thiemann19}, which
implemented an interpreter for a more realistic core functional session typed calculus including recursion and
session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
simple interface to a scheduler. Since our objective was to formalize the list-machine benchmark in Agda and it
does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
of linear and session types as in~\cite{Rouvoet20,Thiemann19}. A formalization of System F$_{\omega\mu}$ was
the subject of \citet{ChapmanKNW19} work, which used an intrinsically-typed representation to
implement a normalization by evaluation for this calculus. The reason behind such formalization effort was
the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$. As our work,
Chapman's et al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
which avoids completely stuck states.}

@@

This paper shows how the combination of intrinsically-typed syntax and definitional interpreters can be used to simplify
the tasks on the formalization of programming languages. Using such approach, we were able to provide a machine-checked version of the
list-machine benchmark in Agda, showing that the approach is useful to formalize both high-level and low-level languages.
The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM,
since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing our work with the
conventional non-dependently typed formalization strategies (like the one used by \citet{Appel07} in their Coq and Twelf implementations),
we can affirm that it requires a fewer number of lines to achieve the same result, even without the use of proof automation.
This happens because the approach uses the power of the host language, and provides some proofs for free due to the
intrinsically-typed syntax.

>>>>

In this work, we extend our previous SBLP 2020 paper~\cite{Feitosa2020} by including support to
indirect jumps, as specified in the version 2.0 of the list
machine benchmark~\cite{AppelDL12}. We provide detailed explanations of the needed
changes on the machine syntax, type system and interpreter to accomodate this extension, which
allows the handling of more complex control flow structures, like the call/return model of function calls.

More specifically, we contribute:

\begin{itemize}
  \item We show how all the details of the list-machine type system
        can be encoded in a dependently-typed syntax setting, which avoids, by construction,
        the presence of stuck states in its definitional interpreter.
  \item We provide a provably correct implementation for testing the subtyping
        relation and to calculate the least common supertype of two input
        types for the machine registers.
  \item We provide a correct-by-construction implementation of a type-checker for
        list-machine programs. Our type-checker produces, as a result, an intrinsically-typed
        representation of the machine code.
  \item We compare the results of an intrinsic approach encoded in Agda with extrinsic
        formalizations encoded in Twelf and Coq. We show that such intrinsic encoding avoids
        unnecessary repetitions and provides some properties for free.
  \item We provide a detailed discussion on the necessary modifications to the type system,
        type checker and interpreter
        to support indirect jumps.
\end{itemize}

@@

Appel's list machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics.
We stick to our initial proposal based on dependently typed syntax and definitional interpreters. Appel's formalization for
the version 2.0 of the benchmark demanded around 1750 LOC without considering his library for the
``very modal model''~\cite{Appel07}. Our formalization demanded 1134 LOC and it depends only on the Agda standard library.
Compared to our solution to the first version of the benchmark (which was implemented in 612 LOC), the inclusion of
indirect jumps increased the formalization size by around 85\%.

@@

\paragraph{Benchmarks for PL mechanization}{
The importance of benchmarks is recognized in several areas in computer science. A relevant example is the
TPTP library for the theorem proving community~\cite{Sut17}. In the context of the programming languages community,
we have the POPLMark challenge~\cite{Aydemir05}, which was developed by a renowned group of programming language researchers aiming
at the collaboration between the PL community and the proof assistants researchers. The main objective of this challenge
was to motivate authors to formalize all of their theorems using such tools. Since the focus of the POPLMark challenge
was mainly the formalization of type soundness theorems, other benchmarks were proposed with different objectives. The list
machine benchmark was proposed by~\cite{Appel07} as an exercise in formalizing results that interest
compiler oriented research and also provides Twelf and Coq solutions to this benchmark. Our work provides another
solution to this benchmark using an intrinsically-typed approach in the Agda programming language. Representation of binding syntax
was the subject of~\cite{FeltyMP18} which proposed a set of problems and research questions for tools
that use the high-order abstract syntax approach for name binding. Since the list-machine benchmark definition avoids
name binding issues because they are orthogonal to most of compiler related proofs~\cite{Appel07}, we just ensure
the correct manipulation of names by following the traditional \emph{de Bruijn} approach. Finally, a recent problem set
was proposed by~\cite{Pientka18}, named POPLMark challenge reloaded, focusing on the mechanization
of logical relation arguments, like strong normalization theorems.}

\paragraph{Definitional interpreters}{
The use of definitional interpreters for specifying semantics dates back to Reynold's pioneer work~\cite{Reynolds72}.
Recently, the interest on such interpreters was revitalized by~\cite{Amin17}, which used definitional
interpreters, implemented in Coq, to prove type soundness theorems for non-trivial typed languages like System F$_{<:}$.
Unlike our work, Amin and Rompf's formalizations do not use intrinsically-typed representations of their syntax, what cluttered
their formalization with the need of dealing with ``stuck states'' in the semantics. \cite{Poulsen18} described how
definitional interpreters for imperative languages can be much concisely implemented by using intrinsically-typed syntax and a
library for name binding using \emph{scope graphs}, which greatly simplifies the treatment of complex
binding structures. Since the list-machine benchmark was designed to address other problems than binding, our representation
using \emph{de Bruijn} indices was sufficient to implement the desired type-checker and interpreter.
Other recent application of definitional interpreters was proposed by~\cite{Rouvoet20}. The main contribution
of Rouvoet et. al. was to define interpreters for linear typed languages supporting session based concurrency. In order to model
linear typing features in the interpreter, the authors have implemented monads based on a separation algebra which they
named \emph{Market}, supporting the main operations for accessing and updating the store used by the interpreter.
Another work using intrinsically-typed syntax for resource control was developed by~\cite{Thiemann19}, which
implemented an interpreter for a more realistic core functional session typed calculus including recursion and
session subtyping in Agda. Thiemann modeled the semantics as an interruptible abstract machine which provides a
simple interface to a scheduler. Since our objective was to formalize the list-machine benchmark in Agda and it
does not have linearity constraints in its state manipulation, we do not need to deal with the complexities
of linear and session types as in~\cite{Rouvoet20,Thiemann19}. A formalization of System F$_{\omega\mu}$ was
the subject of \cite{ChapmanKNW19} work, which used an intrinsically-typed representation to
implement a normalization by evaluation for this calculus. The reason behind such formalization effort was
the verification of a core language for smart-contracts which is based on System F$_{\omega\mu}$. As our work,
Chapman's et al. formalization is an example of how intrisincally-typed syntax leads to clear interpreter code
which avoids completely stuck states.}

@@

This paper shows how the combination of intrinsically-typed syntax and definitional interpreters can be used to simplify
the tasks on the formalization of programming languages. Using such approach, we were able to provide a machine-checked
version of the two versions of the list-machine benchmark in Agda, showing that the approach is useful to
formalize both high-level and low-level languages.
The ideas presented here can be exploited on the formalization of real-world virtual machines, such as the JVM and LUA VM,
since we were able to encode features such as jumps, mutable state, and sub-typing. When comparing our work with the
conventional non-dependently typed formalization strategies (like the one used in~\cite{Appel07} in their Coq and Twelf implementations),
we can affirm that it requires a fewer number of lines to achieve the same result, even without the use of proof automation.
This happens because the approach uses the power of the host language, and provides some proofs for free due to the
intrinsically-typed syntax.
